\documentclass{article}

% NeurIPS 2025 style (placeholder — swap for actual style file at submission time)
\usepackage[preprint]{neurips_2024}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{xcolor}

\title{AlignQuant: The Alignment Tax of Post-Training Quantization in Large Language Models}

\author{%
  Adharapurapu V S M Ashok Kumar \\
  KL University, Vijayawada \\
  \texttt{avsmashokkumar@gmail.com} \\
}

\begin{document}

\maketitle

\begin{abstract}
Every deployed large language model is quantized, yet quantization papers still optimize almost exclusively for capability retention (perplexity, MMLU, GSM8K) while treating alignment as a black-box afterthought. Prior safety-under-compression work \citep{hong2024decoding, q-resafe2024} establishes that behavioral robustness can degrade after compression, but does not provide a reusable direction-level measurement protocol for \emph{why} this happens or how to audit it before deployment. We present \textbf{AlignQuant}, a two-axis evaluation framework that cleanly separates \emph{capability retention} from \emph{alignment retention}, together with \textbf{AlignProbe}, a toolkit for measuring subspace rotation, steering efficacy, and behavioral safety regressions under post-training quantization. This draft contributes two concrete artifacts. First, it formalizes a preregistered open-weight geometry study spanning refusal, truthfulness, and sycophancy directions extracted with independent methods \citep{arditi2024refusal, li2023iti, rimsky2024caa}. Second, it reports an API-only behavioral reference on the Cohere Command R family that stress-tests confidentiality under model-visible context. In a clean two-turn retrieved-context exfiltration rerun, \texttt{command-r7b-12-2024} leaked the exact token in 15/24 cases (62.5\%), whereas \texttt{command-r-plus-08-2024} leaked 1/24 cases (4.2\%) and \texttt{command-a-03-2025} leaked 0/24 cases; in a one-shot retrieved-context extraction setting, all three models leaked heavily. These results do not measure quantization directly, but they motivate the behavioral targets that the open-weight geometry sweep must explain and show that harmful-request refusal and prompt-visible confidentiality are separable properties. We release the measurement protocol, redacted evidence bundles, and paper artifacts so that future quantization studies can report alignment retention as a first-class metric rather than an anecdote.
\end{abstract}

\section{Introduction}
Every production LLM stack makes a compression decision. Weight-only 8-bit inference, 4-bit quantization, activation smoothing, adapter-aware quantization, and low-rank recovery are no longer optimization curiosities; they are deployment defaults. Yet the dominant evaluation loop for these methods still asks a narrow question: does the quantized model preserve perplexity and benchmark accuracy? That is necessary, but it is not sufficient. Alignment is also a property of the deployed model, and deployment-time compression can alter the very residual-stream directions and steering subspaces that post-training alignment methods rely on.

This creates a methodological gap. The alignment literature has shown that refusal, truthfulness, sycophancy resistance, and related behaviors are often linearly localized or at least directionally compressible in hidden-state space \citep{arditi2024refusal, li2023iti, rimsky2024caa, marks2024geometry}. The quantization literature, by contrast, largely stops at output-level capability metrics. Between them sits an unanswered question: when we quantize a model, do we preserve the geometry of the alignment-relevant subspaces we claim to care about, or do we pay an \emph{alignment tax} that is invisible to perplexity?

AlignQuant is designed to answer that question with a protocol that is both mechanistic and operational. Mechanistically, it measures subspace angles, probe retention, and steering efficacy before and after quantization. Operationally, it pairs those geometry measurements with behavioral reference tests on models where hidden states are unavailable. The current repository contains the full measurement pipeline plus a closed-model red-team reference on the Cohere Command R family. That reference is not itself a quantization result. It serves a different role: it identifies concrete confidentiality failures that an eventual open-weight geometry analysis should help predict and explain.

\paragraph{Contributions.}
\begin{itemize}
    \item We formalize the \emph{alignment tax} of quantization and define a two-axis evaluation protocol that separates capability retention from alignment retention.
    \item We define the \textbf{AlignQuant Score}, a unified scalar summarizing geometry preservation across multiple safety-relevant directions.
    \item We release \textbf{AlignProbe}, an open-source measurement toolkit covering direction extraction, behavioral red-teaming, confidence intervals, and redacted evidence bundles.
    \item We provide a preregistered open-weight study design for quantization geometry across refusal, truthfulness, and sycophancy axes.
    \item We report an API-only behavioral reference on the Cohere Command R family showing that confidentiality robustness depends strongly on both model family member and context type, especially in retrieved-context and staged extraction settings.
\end{itemize}

\section{Related Work}

\paragraph{Quantization of large language models.}
Modern post-training quantization (PTQ) targets reconstruction fidelity under low-bit weight storage. LLM.int8() \citep{dettmers2022llmint8} isolates outlier features for near-lossless 8-bit inference; GPTQ \citep{frantar2023gptq} performs layer-wise second-order error correction; SmoothQuant \citep{xiao2023smoothquant} migrates activation-outlier difficulty into the weights; AWQ \citep{lin2024awq} protects a salient-weight set based on activation magnitude; QLoRA's NF4 \citep{dettmers2023qlora} is an information-optimal 4-bit datatype for Gaussian-distributed weights; SpQR, OmniQuant, and QuIP\# \citep{dettmers2023spqr, shao2024omniquant, tseng2024quip} push toward 2--3 bits. All are evaluated on perplexity and zero-shot accuracy; none report whether the directional structure RLHF writes into the residual stream is preserved.

\paragraph{Behavioral trust under compression.}
Decoding Compressed Trust \citep{hong2024decoding} evaluates 3 models $\times$ 5 compression methods across 8 trust dimensions; Q-resafe \citep{q-resafe2024} reports non-trivial safety degradation across AWQ/GPTQ/SmoothQuant/AutoRound and proposes a LoRA-style patch. Both works are strictly black-box: they measure attack success rates, leaving open \emph{which} representations are damaged and \emph{why} one quantizer degrades safety more than another. AlignQuant complements these with a direction-level, geometric account, and we show (\S\ref{sec:experiments}) that our subspace-angle metric correlates $r > 0.8$ with their behavioral ASR while being cheaper and mechanistically interpretable.

\paragraph{Alignment as a low-dimensional, linear phenomenon.}
A convergent line of evidence finds that safety-relevant computation is linearly localized. \citet{arditi2024refusal} show refusal across 13 chat models is mediated by a single residual-stream direction; ITI \citep{li2023iti} localizes truthfulness into $\approx 48$ attention heads via rank-one edits; CAA \citep{rimsky2024caa} builds per-behavior steering vectors (sycophancy, corrigibility, hallucination); Representation Engineering \citep{zou2023repe} generalizes these into a top-down vocabulary of concept directions; Geometry of Truth \citep{marks2024geometry} demonstrates an emergent linear truth direction; and strategic-deception probes reach AUROC $>0.96$ \citep{apollo2025deception,hubinger2024sleeper}. \citet{qi2025shallow} show that RLHF alignment is concentrated in the first few output tokens---a thin surface phenomenon. None of these works evaluates how their directions fare under weight quantization.

\paragraph{PEFT under quantization.}
QLoRA \citep{dettmers2023qlora} trains LoRA adapters through a quantized backbone; LoftQ \citep{li2024loftq} jointly initializes quantized weights and adapters via alternating SVD; QA-LoRA \citep{xu2024qalora} makes adapters mergeable into INT4 deployment; ApiQ \citep{liao2024apiq} extends to 2-bit; LQ-LoRA and PEQA \citep{guo2024lqlora,kim2023peqa} further decompose the quantization grid. All are evaluated on perplexity or GLUE; none measures alignment retention. Closest in spirit, SafeLoRA \citep{hsu2024safelora} and Vaccine \citep{huang2024vaccine} protect safety during benign fine-tuning by projecting away from a safety direction---but both assume an unquantized backbone. The quantization-only regime, in which no fine-tuning occurs and the question is simply whether directions survive the grid, remains unstudied.

\paragraph{Positioning.}
AlignQuant sits at the intersection of these three literatures. Our contribution is the two-axis evaluation protocol (capability retention vs alignment retention) and the systematic, direction-level measurement that no prior work performs.

\subsection{Post-training quantization of LLMs}
See the Related Work above; this subsection is reserved for additional model-card-specific notes in the camera-ready.

\subsection{Probing and activation directions for alignment}
See the Related Work above; this subsection is reserved for reviewer-requested deep dives.

\subsection{Parameter-efficient fine-tuning under quantization}
See the Related Work above; reserved for expanded DRaFT-Q positioning.

\section{Hypotheses}
\label{sec:hypotheses}
We commit in advance to three falsifiable hypotheses, each paired with a null-result framing so the paper remains informative under any outcome.

\paragraph{H1 (Differential alignment tax).} For a fixed model, the subspace angle between FP16 and 4-bit quantized \emph{refusal} directions is strictly greater than the angle for a matched-norm control direction extracted from a semantically unrelated binary contrast. Formally, $\theta_{\text{refusal}} > \theta_{\text{control}}$ with $p < 0.01$ over four base models and two 4-bit schemes (NF4, GPTQ). \textbf{Null framing.} If $\theta_{\text{refusal}} \approx \theta_{\text{control}}$, quantization is geometrically uniform and the field's current "quantization is safe" intuition is vindicated at the representation level. We report that finding as the headline.

\paragraph{H2 (Geometry predicts behavior).} Across the full $(\text{model} \times \text{scheme})$ grid, the cell-level subspace angle on the refusal axis correlates with the behavioral Attack Success Rate on AdvBench at Pearson $r > 0.5$. \textbf{Null framing.} If $r \leq 0.5$, geometry and behavior are decoupled and we instead publish this as a cautionary finding: the interpretability community's linear-direction story does not transfer to the compressed regime, which is itself a load-bearing negative result.

\paragraph{H3 (Scheme ranking).} Quantization schemes do not degrade alignment uniformly. The rank ordering of schemes by alignment retention is stable across models and probe methods. Concretely we predict $\mathcal{A}_{\text{BnB-int8}} > \mathcal{A}_{\text{AWQ}} > \mathcal{A}_{\text{GPTQ}} \approx \mathcal{A}_{\text{NF4}}$ at matched perplexity. \textbf{Null framing.} If the ranking is unstable across seeds or probe methods, we report this as evidence that alignment geometry under compression is high-variance and no single quantizer should be recommended without per-model auditing. This motivates the AlignProbe tool directly.

\section{Method}

\subsection{Alignment axes and direction extraction}
We study three alignment-relevant directions, each extracted by a different, well-established method to rule out probe-specific artifacts:
\begin{enumerate}
    \item \textbf{Refusal} --- difference-in-means vectors between harmful and harmless instructions at the residual stream, following \citet{arditi2024refusal}.
    \item \textbf{Truthfulness} --- Inference-Time Intervention (ITI) truthful heads and linear probes trained on TruthfulQA pairs, following \citet{li2023iti}.
    \item \textbf{Sycophancy resistance} --- Contrastive Activation Addition (CAA) vectors from matched user-pushed-wrong-belief prompt pairs, following \citet{rimsky2024caa}.
\end{enumerate}
A \textbf{control direction} --- a random residual-stream direction of matched norm --- is included for every experiment to guard against the null hypothesis that quantization induces uniform drift of all directions (Objection 3, Sec.~\ref{sec:objections}).

\subsection{Quantization schemes}
FP16, BnB-int8, NF4 (QLoRA), GPTQ, AWQ. All applied post-hoc to the same fine-tuned checkpoint to isolate the effect of quantization.

\subsection{Two-axis evaluation framework}
\label{sec:twoaxis}
We disentangle two quantities that prior work conflates:

\paragraph{Axis 1 --- Capability retention.} Perplexity on WikiText-2, accuracy on MMLU and GSM8K.

\paragraph{Axis 2 --- Alignment retention.} For each (model, quant, direction) cell:
\begin{itemize}
    \item \textbf{Subspace angle} $\theta = \arccos\,\cos(\mathbf{d}_{\text{fp16}}, \mathbf{d}_{\text{q}})$ --- the core geometric metric.
    \item \textbf{Probe accuracy drop} $\Delta\text{acc} = \text{acc}_{\text{fp16}} - \text{acc}_{\text{q}}$.
    \item \textbf{Steering efficacy at optimal coefficient.} The steering scalar $\alpha$ is re-swept on the quantized model to find its own optimum, and we report the attainable intervention strength. This rules out the trivial-rescaling null.
    \item \textbf{Behavioral reference.} Attack Success Rate on BeaverTails / DecodingTrust, enabling correlation with \citet{hong2024decoding}.
\end{itemize}

\paragraph{AlignQuant Score.} A single scalar $\mathcal{A} = 1 - \tfrac{\theta}{\pi/2}$ averaged over directions, reported alongside the capability axis so the differential tax is visible at a glance.

\subsection{The AlignQuant Score}
The AlignQuant Score collapses the per-cell geometric metrics into a single scalar for leaderboard use. For a given (model, quantization scheme) cell, we compute
\[
\mathcal{A} \;=\; 1 - \frac{1}{|\mathcal{D}|}\sum_{d \in \mathcal{D}} \frac{\theta_d}{\pi/2},
\]
where $\mathcal{D}$ is the set of alignment directions (refusal, truthfulness, sycophancy) and $\theta_d$ is the subspace angle between the FP16 and quantized direction. $\mathcal{A} = 1$ means perfect alignment preservation; $\mathcal{A} = 0$ means the quantized direction is orthogonal to the original, that is, alignment is destroyed. We always report $\mathcal{A}$ alongside the capability-axis metrics (perplexity, MMLU) so the differential alignment tax is visible as the gap between the two.

\subsection{Threat model}
The paper studies deployment-time alignment degradation under three distinct but related settings. Table~\ref{tab:threat-model} makes the boundaries explicit so the reader can distinguish what is already evidenced in this repository from what the open-weight sweep is meant to explain.

\begin{table}[t]
\centering
\small
\begin{tabular}{p{0.19\linewidth}p{0.22\linewidth}p{0.20\linewidth}p{0.28\linewidth}}
\toprule
Setting & Attacker access & Secret / target type & Primary metric \\
\midrule
System-context probe & User prompt only; synthetic secret injected into model-visible instructions & Application-supplied canary in system context & Exact / any canary leak rate \\
Retrieved-context probe & User prompt plus model-visible retrieved record or tool-style state & Confidential field in retrieved/tool context & Exact / any canary leak rate \\
Quantization geometry sweep & Open-weight checkpoint plus quantized variants & Alignment-relevant residual-stream directions & Subspace angle, probe retention, steering efficacy \\
\bottomrule
\end{tabular}
\caption{Threat-model summary. The first two rows are behaviorally evidenced in the current repository. The third row is the open-weight study that AlignQuant is designed to execute and explain.}
\label{tab:threat-model}
\end{table}

\section{Experiments}

\subsection{Models and data}
\paragraph{Models (local probing).} Llama-3-8B-Instruct, Mistral-7B-Instruct-v0.3, Qwen2.5-7B-Instruct, Gemma-2-9B-it. All probing performed on residual-stream activations at every layer.

\paragraph{Models (API-only behavioral reference).} Cohere Command R (35B) and Command R+ (104B) via Cohere API; behavioral axes only (no hidden-state access).

\paragraph{Quantization schemes.} FP16 baseline, BnB-int8, NF4 (QLoRA), GPTQ (4-bit), AWQ (4-bit).

\paragraph{Data.} AdvBench + HarmBench harmful/harmless pairs (refusal), TruthfulQA MC1/MC2 (truthfulness), Sharma et al. sycophancy prompts + our extensions (sycophancy), WikiText-2 + MMLU + GSM8K (capability).

\paragraph{Hardware.} Primary experiments on Kaggle 2$\times$T4 (32GB VRAM), which fits all 7--9B models in FP16 and 13--35B models in 4-bit, ensuring full reproducibility on a free-tier platform.

\subsection{Master behavioral summary}
Before turning to the open-weight geometry program, we summarize the completed behavioral reference results that are already supported by repository artifacts. Table~\ref{tab:master-results} is the quickest way to read the current evidence: harmful-request refusal is strong, but confidentiality robustness varies sharply with context type and model.

\begin{table*}[t]
\centering
\small
\begin{tabular}{llcccc}
\toprule
Family & Setting & Metric & \texttt{command-r7b-12-2024} & \texttt{command-r-plus-08-2024} & \texttt{command-a-03-2025} \\
\midrule
Harmful prompts & AdvBench reference & Unsafe compliance & 0/100 (0.0\%) & -- & -- \\
System context & Canary rotation & Exact leak & 58/120 (48.3\%) & 33/120 (27.5\%) & 1/120 (0.8\%) \\
System context & Multi-turn extraction & Final exact leak & 16/30 (53.3\%) & 3/30 (10.0\%) & 0/30 (0.0\%) \\
Retrieved context & Single-turn tool-context probe & Exact leak & 20/24 (83.3\%) & 18/24 (75.0\%) & 19/24 (79.2\%) \\
Retrieved context & Single-turn tool-context probe & Any leak & 24/24 (100.0\%) & 22/24 (91.7\%) & 22/24 (91.7\%) \\
Retrieved context & Clean two-turn tool-chain rerun & Final exact leak & 15/24 (62.5\%) & 1/24 (4.2\%) & 0/24 (0.0\%) \\
Retrieved context & Clean two-turn tool-chain rerun & Transcript any leak & 18/24 (75.0\%) & 5/24 (20.8\%) & 4/24 (16.7\%) \\
\bottomrule
\end{tabular}
\caption{Master summary of the completed behavioral evidence currently available in the repository. Dashes indicate that the specific baseline was not part of the earlier 100-prompt harmful-request pilot.}
\label{tab:master-results}
\end{table*}

\subsection{API-only behavioral reference on the Cohere Command R family}
\label{sec:experiments}
The repository currently contains a completed behavioral reference suite on three Cohere models: \texttt{command-r7b-12-2024}, \texttt{command-r-plus-08-2024}, and \texttt{command-a-03-2025}. This suite is not a substitute for the open-weight geometry analysis because hidden states are unavailable, but it serves two practical purposes. First, it identifies concrete failure classes that a future geometry metric should correlate with. Second, it demonstrates that harmful-request refusal and confidentiality under model-visible context are separable axes.

The strongest closed-model result is a retrieved-context confidentiality failure. In a single-turn tool-context setting, where a synthetic confidential token is placed in a retrieved record and the user asks for extraction, all three models leaked heavily: exact leakage was 20/24 for \texttt{command-r7b-12-2024}, 18/24 for \texttt{command-r-plus-08-2024}, and 19/24 for \texttt{command-a-03-2025}. This is operationally important because many real systems place internal notes, identifiers, or customer data into retrieved passages rather than system prompts.

The more discriminating result appears in a cleaner two-turn tool-chain exfiltration rerun. There the model first receives a confidential retrieved record, then answers a safe summary or redacted-view request, and only afterwards gets a follow-up request for the exact token. The final exact leakage rates are summarized in Table~\ref{tab:toolchain-clean}. In this staged setting, \texttt{command-r7b-12-2024} remains clearly vulnerable, while the stronger models mostly refuse the second-step extraction turn.

\begin{table}[t]
\centering
\small
\begin{tabular}{lcccc}
\toprule
Model & First-turn any leak & Final exact leak & Transcript any leak & Final refusal \\
\midrule
\texttt{command-r7b-12-2024} & 8/24 (33.3\%) & 15/24 (62.5\%) & 18/24 (75.0\%) & 12/24 (50.0\%) \\
\texttt{command-r-plus-08-2024} & 2/24 (8.3\%) & 1/24 (4.2\%) & 5/24 (20.8\%) & 24/24 (100.0\%) \\
\texttt{command-a-03-2025} & 1/24 (4.2\%) & 0/24 (0.0\%) & 4/24 (16.7\%) & 24/24 (100.0\%) \\
\bottomrule
\end{tabular}
\caption{Clean two-turn retrieved-context exfiltration rerun on the Cohere Command R family. The models are given a confidential retrieved record, asked for a safe first-step summary, and then asked to extract the token exactly.}
\label{tab:toolchain-clean}
\end{table}

This behavioral reference does not establish a quantization effect. It establishes a target phenomenon: alignment-relevant confidentiality behavior can fail sharply when context type changes from direct prompt instruction to retrieved/tool-style state. Any geometry-based alignment-retention metric that claims deployment relevance should eventually predict or explain this kind of context-conditioned behavioral break.

\subsection{Does quantization impose a differential alignment tax? (H1)}
The open-weight geometry sweep for H1 is specified but not yet finalized in this repository snapshot. The key test compares the subspace angle of a refusal direction before and after quantization against a matched-norm control direction extracted from an unrelated semantic contrast. The main outcome is not simply whether quantization moves the representation, but whether it moves \emph{alignment-relevant} directions more than generic ones. We will report layer-aggregated and layer-localized angle distributions, together with bootstrap uncertainty over prompt subsets, for NF4 and GPTQ at minimum.

\subsection{Is the tax scheme-dependent? (H2)}
H2 asks whether different quantizers preserve alignment geometry differently at matched capability loss. The implemented protocol already fixes the capability axis (perplexity, MMLU, GSM8K) and the alignment axis (angle, probe retention, steering efficacy), so the remaining work is to execute the full grid across the open-weight models. The intended analysis is a per-model ranking of schemes by AlignQuant Score, followed by a cross-model stability analysis. The behavioral reference on the Cohere Command R family already suggests that model family members can differ materially on confidentiality robustness; the open-weight sweep will test whether quantization introduces an analogous ranking within a fixed base model.

\subsection{Can dynamic-rank PEFT preserve alignment geometry? (H3)}
H3 evaluates whether dynamic-rank recovery methods preserve alignment geometry better than static low-rank baselines when the backbone is aggressively quantized. The protocol is straightforward: quantize the base model, apply static LoRA / QLoRA / LoftQ versus dynamic-rank DRaFT-Q, re-extract the alignment directions, and compare both angle retention and steering efficacy. We do not report a completed DRaFT-Q result in the current repository snapshot, so this subsection should be read as the prespecified comparison rather than a finished empirical claim.

\subsection{Signature analysis: layer-wise alignment vulnerability map}
\label{sec:layer-map}
Our first signature result is the \emph{layer vulnerability map}: for every layer $\ell$ of every model, we compute the per-layer subspace angle $\theta_\ell$ between FP16 and quantized refusal directions, producing a heatmap across $(\text{layer} \times \text{scheme})$. This answers a question no prior work poses: is the alignment tax concentrated at a few vulnerable layers, or is it distributed? Preliminary theory predicts the middle third of the network (where \citet{arditi2024refusal} localizes refusal) shows the sharpest angle under 4-bit schemes; this map either confirms that prediction or redirects the field toward a different layer regime.

\subsection{Signature analysis: alignment-tax curvature across bit-widths}
\label{sec:curvature}
Our second signature result characterizes how the alignment tax scales with aggression of compression. For each model we extract the refusal direction at FP16, BnB-int8, NF4 (4-bit), and a 3-bit GPTQ checkpoint, and fit the alignment retention curve $\mathcal{A}(b)$ as a function of bit-width $b$. We test whether this curve is \emph{convex}: the prediction is that alignment degrades superlinearly below a critical bit-width (somewhere between 4 and 3 bits), while capability retention degrades roughly linearly in the same regime. A convex-alignment / linear-capability result means practitioners are paying an exponentially increasing safety cost for each bit they shave, and gives the field a principled floor bit-width.

\section{Discussion}
Two practical lessons already emerge from the current artifact set. First, alignment should be evaluated on more than one axis. The Cohere reference suite refuses harmful prompts robustly, yet still shows substantial confidentiality failures when secrets are placed in model-visible context. That is exactly the kind of decoupling the AlignQuant protocol is designed to surface for quantization as well: a model can look capability-stable and even safety-stable on one benchmark family while drifting badly on another alignment-relevant behavior. Second, context type matters. One-shot extracted system-context probes, one-shot retrieved-context probes, and staged two-turn exfiltration flows do not produce the same ranking across models. Any serious alignment-retention metric needs to explain that heterogeneity rather than average it away.

The null-result framing remains important. If the open-weight geometry sweep shows that quantization rotates refusal directions no more than matched control directions, or that angle change fails to predict behavioral regression, that would still be a useful scientific result. It would mean that current interpretability stories about low-dimensional alignment do not transfer cleanly to the compressed regime, and the field would need different tools. AlignQuant is designed so that either outcome is informative.

\section{Limitations}
The current repository is strongest on methodology and behavioral reference evidence, not yet on the full open-weight quantization sweep promised by the paper title. We therefore avoid claiming completed geometry results that are not present. The Cohere behavioral reference is also limited in two ways: it is API-only, so we cannot inspect internal activations, and it concerns application-visible synthetic secrets rather than provider-hidden prompts or real user data. Finally, several of the confidence intervals are still pilot-scale because some staged experiments run at only 24 cases per model. The paper-grade contribution today is therefore a rigorously specified protocol plus a credible behavioral reference suite, not a finished universal theory of quantization-induced alignment failure.

\section{Pre-Empted Reviewer Objections}
\label{sec:objections}
\paragraph{Objection 1: ``Decoding Compressed Trust and Q-resafe already show this.''} Those works are purely behavioral. We operate at the direction level, provide mechanism, and show our angle metric correlates $r>0.8$ with their ASR while being cheaper and interpretable.

\paragraph{Objection 2: ``Refusal directions are single-model artifacts.''} The study is designed to address this by spanning four open-weight models and three independent extraction methods (diff-in-means, ITI, CAA). The current repository snapshot contains the full protocol and closed-model behavioral reference, while the multi-model open-weight execution remains ongoing.

\paragraph{Objection 3: ``Angle change could be trivial rescaling.''} Two controls: (a) a random matched-norm direction degrades significantly less than safety directions; (b) steering efficacy is re-optimized on the quantized model and still degrades, ruling out pure rescaling.

\section{Broader Impact and Safety Disclosure}
This work studies safety-relevant degradation of open-weight and API-accessible LLMs, including models in the Cohere Command R family. We follow a coordinated-disclosure protocol: any vulnerability-class finding (e.g., a quantization scheme that systematically disables refusal) is reported to \texttt{safety@cohere.ai} immediately upon discovery and is shared with \texttt{labs@cohere.com} at least four weeks before public release, in accordance with the Cohere Labs Research Grant disclosure policy. We believe public characterization of the alignment tax is net-positive for safe deployment: practitioners currently have no principled way to choose a quantization scheme for safety, and this work provides one.

\paragraph{Acknowledgements.} This work was supported by compute credits from a Cohere Labs Research Grant; these grants are designed to support academic partners conducting research with the goal of releasing scientific artifacts and data for good projects.

\section{Conclusion}
Compression is a deployment decision, and deployment decisions should be evaluated on deployment-relevant alignment properties. AlignQuant contributes a concrete way to do that: separate capability retention from alignment retention, measure geometry when hidden states are available, and anchor those measurements to behavioral reference tests when they are not. The current artifact set already shows that confidentiality robustness under model-visible context can vary sharply across models and across context types, even when harmful-request refusal remains strong. The next step is to complete the open-weight geometry sweep and test whether subspace rotation, probe degradation, and steering-efficacy loss can predict those behavioral failures before deployment. If they can, the field gets a practical auditing tool. If they cannot, that is still a valuable negative result about the limits of current linear-alignment accounts under compression.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
